---
title: "Rcpp Tutorial"
author: "Marvin N. Wright"
date: "Aug 1, 2017"
output: 
  ioslides_presentation: 
    transition: faster
    css: smallfont.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Why Rcpp?

## Sometimes R is very slow

* Random element-wise access
* Loops that cannot be vectorized
* Recursive functions

## Sometimes R is very slow

Fibonacci sequence
```{r}
fibR <- function(n) {
  if (n < 2) {
    return(1)
  } else {
    return(fibR(n - 1) + fibR(n - 2))
  }
}
```

```{r}
system.time(fibR(40))
```

<div class="footnotes">Example from http://gallery.rcpp.org/articles/fibonacci-sequence/.</div>

# How to use Rcpp?

## Inline C++ code

```{r}
library(Rcpp)

cppFunction("double mysum(double a, double b) {
   return a + b;         
}")

mysum(3.5, 2.1)
```

## Inline C++ code (using vectors)

```{r}
cppFunction("double myvecsum(Rcpp::NumericVector v) {
  double result = 0;
  for (auto& x : v) {
    result += x;
  }
  return result;
}")

myvecsum(c(3.5, 2.1, 4.6))
```

## C++ files

Save to `mysum.cpp`:
```{Rcpp, eval=FALSE}
#include <Rcpp.h>

// [[Rcpp::export]]
double mysum(double a, double b) {
   return a + b;         
}
```

From R:
```{r}
sourceCpp("mysum.cpp")
mysum(3, 4)
```

## Rcpp in R Markdown

Just use:

    ```{Rcpp}`r ''`
    // [[Rcpp::export]]
    double mysum5(double a, double b) {
      return a + b;         
    }
    ```

## Rcpp Sugar

* Rcpp provides typical R functions in C++
* Examples: `min()`, `max()`, `sum()`, `mean()`, `rnorm()`, `pnorm()`, `dnorm()`, `qnorm()`, etc.


```{r}
cppFunction("Rcpp::NumericVector myexp(Rcpp::NumericVector vec) {
  return Rcpp::exp(vec);
}")

myexp(c(5, 4, 7))
```

## Rcpp Sugar 

Rcpp uses the R random number generator
```{r}
cppFunction("Rcpp::NumericVector myrnorm(int n) {
  return Rcpp::rnorm(n);
}")

set.seed(10)
rnorm(5)

set.seed(10)
myrnorm(5)
```

## Rcpp data types

|Value | R vector|Rcpp vector|Rcpp matrix|C++ scalar|
|:---:|:---:|:---:|:---:|:---:|
|Logical|`logical`  |`LogicalVector`| `LogicalMatrix`|`bool`|
|Integer|`integer`  |`IntegerVector`|`IntegerMatrix`|`int`|
|Real|`numeric` |`NumericVector`|`NumericMatrix`|`double`|

<div class="footnotes">Complete table at https://teuder.github.io/rcpp4everyone_en/.</div>

# How fast is Rcpp? 

## Back to Fibonacci numbers

```{r}
fibR <- function(n) {
  if (n < 2) {
    return(1)
  } else {
    return(fibR(n - 1) + fibR(n - 2))
  }
}
```

```{Rcpp}
// [[Rcpp::export]]
int fibCpp(int n) {
  if (n < 2) {
    return 1;
  } else {
    return fibCpp(n - 1) + fibCpp(n - 2);
  }
}
```

## Back to Fibonacci numbers

```{r}
library(microbenchmark)
microbenchmark(R = fibR(40), 
               Cpp = fibCpp(40), 
               times = 1, unit = "s")
```

# How to use Rcpp in a package? 

## Live demonstration

# Final remarks

## Discussion

### Advantages
* Fast and memory efficient (potentially)
* Object-oriented
* Great libraries available, e.g. Eigen, Armadillo

### Disadvantages
* More work
* Code less readable
* More bugs?
* Not always faster! 

## Installation

1. System preparation
    * Linux: Install R development package, .e.g., `r-devel` or `r-base-dev`
    * Windows: Install RTools from https://cran.r-project.org/bin/windows/Rtools/
    * Mac: Install Xcode command line tools, i.e., `xcode-select --install`
2. Install `Rcpp` package: `install.packages("Rcpp")`

## Resources

* Rcpp website: http://www.rcpp.org
* Rcpp paper in JSS: http://dx.doi.org/10.18637/jss.v040.i08
* Rcpp documentation (technical): http://dirk.eddelbuettel.com/code/rcpp/html/
* Rcpp quick reference: https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-quickref.pdf
* Chapter in Hadley Wickham's book: http://adv-r.had.co.nz/Rcpp.html
* Rcpp gallery: http://gallery.rcpp.org/


